NumPy---Array Programming in Python

### 2020-01-29 body.tex
Clean up
Remove future dtype discussion from main text

% pandas, xarray, ... - leave out Pandas since it's not an array?
% distributed arrays, delayed execution arrays
% GPU arrays, tensor computation libraries
% sparse arrays
% foreign language arrays

% Ralf: we should describe our role at the foundation of the stack
% i was thinking something along these lines ():
NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It supports many types of CPUs (e.g., POWER and ARM) with similar
performance as native C code.
% a brief word about accelerators perhaps?
% end with how this fits the bill for a huge number of users

@@

% Ralf: should mention that there are typically utilities to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill the resultant gaps led to a
proliferation of array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

Importantly, some projects chose to follow the NumPy API exactly---with Dask,
CuPy, Sparse and JAX the most prominent examples---while others have
NumPy-inspired semantics but chose to deviate in places.
% Ralf: may want to add something like:
% when they deviate, they sometimes fix it, e.g.,
% https://github.com/pytorch/pytorch/pull/1983
% https://github.com/pytorch/pytorch/issues/2228
% many have for numpy users
% https://github.com/torch/torch7/wiki/Torch-for-Numpy-users

% \footnote{\url{https://cupy.chainer.org}}

@@

% Ralf: I am not sure where this goes, but it would be good to have something
% about dtype limitation in this section and i also think there should
% be some discussion of missing values somewhere
% (it is mentioned below, but not discussed)
While it supports a wide range of datatypes, it does not have good support
for missing values or categorical data. % may be worth saying a few sentences about missing values
% if we want to discuss Pandas, perhaps we could do so here (i.e, missing values)
% maybe we need to rename the section, but this could anticipate
% the dtypes discussion below

\section*{Interoperability and extensibility}

NumPy supports array operations between external array objects by
acting as a central coordination mechanism with a well-specified API.
With the proliferation of array implementations discussed above, a
widespread adoption of the NumPy API is valuable: it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray, by facilitating
the development of specialized solutions and new tools that operate in
concert.

There has been a steady increase in the number of external libraries that
provide arrays and a NumPy-like API for manipulating them.
Ideally, operating on these arrays using NumPy functions or semantics would
simply work, so that end users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays, and so
forth, as appropriate.

@@

Subclasses of NumPy often exist due to difficulty in constructing richer data
types, such as quantities with physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values. % pandas: categoricals, IP addresses?
To improve \emph{extensibility}, we are currently overhauling the data type
system to make it easier to implement data types both in Python and C, and to
support these and other applications.
The widespread adoption of rich, user defined, data types can allow more
concise, feature rich scientific computing.
By making these user implemented data types a fundamental design of NumPy
these tools will enabled and interact naturally with different libraries.
% need to add a bit of clarity on what "enable and interact naturally with different libraries" means.
%
%  1. I expect multiple user dtypes can interact (naturally); 2. libraries such
%  as sklearn (or dask) might "just work" with e.g. physical Units (and
%  Categoricals, which they probably really would like to use e.g. for labels);
%  3. Things being inside a dtype rather than subclass is a better user
%  experience.

>>>>

NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It supports many types of CPUs (e.g., POWER and ARM) with similar
performance as native C code.
% a brief word about accelerators perhaps?
% end with how this fits the bill for a huge number of users

@@

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill the resultant gaps led to a
proliferation of array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

Importantly, some projects chose to follow the NumPy API exactly---with Dask,
CuPy, Sparse and JAX the most prominent examples---while others have
NumPy-inspired semantics but chose to deviate in places.
% when they deviate, they sometimes fix it, e.g.,
% https://github.com/pytorch/pytorch/pull/1983
% https://github.com/pytorch/pytorch/issues/2228
% many have for numpy users
% https://github.com/torch/torch7/wiki/Torch-for-Numpy-users

@@

Ideally, operating on these arrays using NumPy functions or semantics would
simply work, so that end users could write code once, and would then benefit
from switching between NumPy arrays, GPU arrays, distributed arrays, and so
forth, as appropriate.
To support array operations between external array objects, NumPy
acts as a central coordination mechanism with a well-specified API.
The widespread adoption of the NumPy API is valuable: it lowers the
barrier to entry for newcomers and provides the wider community with a
stable array programming interface. This, in turn, prevents disruptive
schisms like the divergence of Numeric and NumArray, by facilitating
the development of specialized solutions and new tools that operate in
concert.

@@
